Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-01-21 xrb2/paper.tex
Unblue some things that have already been reviewed

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. {\color{blue}We saw in \citet{flame_wave1} that 
increasing the 
neutron star rotation rate reduces the horizontal lengthscale of the flame. An estimate of 
this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius may be 
thought of as the scale over which the balance between the Coriolis force and horizontal pressure 
gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use the energy generation rate $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region (where significant energy generation is occurring) as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). From Equation \ref{eqn:rossby}, we can see that increasing 
the rotation rate from $500~\mathrm{Hz}$ to $1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater 
confinement from the Coriolis force should reduce the horizontal extent of the flame by a similar 
factor. However, the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our simulations we see that this scaling does not work so well for all rotation rates. The simulations seem to follow the theoretical
scaling described in Equation \ref{eqn:rossby} more closely at higher rotation rates ($1000~\mathrm{Hz}$ and 
higher), based on the results of \citet{flame_wave1}.}

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. {\color{blue} The $\dot{e}_\mathrm{nuc}$ plots in Figure \ref{fig:time_series_enuc_1000} also support this interpretation, with the interior of the flame nearest to the crust\MarginPar{What is meant by interior here?? The edge? The part nearest the crust?} in the {\tt F1000} run reaching higher $\dot{e}_\mathrm{nuc}$ values than for the {\tt F500} run in Figure \ref{fig:time_series_enuc_500}.} In contrast to {\tt F500} and {\tt F1000}, the 
lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition for {\tt F250} also 
aligns with the reasoning given above, with the lower rotation in this case potentially leading to 
insufficient confinement such that the temperature and density required for ignition is not 
achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) would need to 
take over at lower rotation rates to allow a burst to occur, at least for the initial model used 
here.

@@

\begin{deluxetable}{lrrr}
	\tablecaption{\label{table:flame_speeds} Flame speed and acceleration values measured for each 
	simulation. $v_0$ is the flame velocity in the $+r$ direction at $t 
	= 0~\mathrm{ms}$. The initial flame 
	velocities and accelerations are derived from a quadratic least-squares fit to each of the 
	datasets for times $t \gtrsim 20~\mathrm{ms}$. {\color{blue} Using these fit parameters we calculate the velocity at $t = 50~\mathrm{ms}$, $v_{50}$, at which point the flames have reached steady propagation.}}
	\tablehead{\colhead{run} & \colhead{$v_0$ (km s$^{-1}$)} & \colhead{$a$ (km s$^{-2}$)} & \colhead{$v_{50}$ (km s$^{-1}$)}} 
	\startdata
  {\tt F1000}     & $3.414 \pm 0.008$ & $-1.03 \pm 0.13$ & $3.36 \pm 0.01$\\
  {\tt F500}      & $3.077 \pm 0.013$ & $5.41 \pm 0.19$ & $3.35 \pm 0.02$\\
  {\tt F500\_2E8} & $3.760 \pm 0.015$ & $1.74 \pm 0.25$ & $3.85 \pm 0.02$\\
  {\tt F500\_3E8} & $-5.293 \pm 0.100$ & $357.41 \pm 2.15$ & $12.58 \pm 0.15$\\
	\enddata
	%
\end{deluxetable}

@@

Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the characteristic flame structure shown in $\bar{A}$ plots for the lower temperature $500~\mathrm{Hz}$ runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. {\color{blue}A similar effect is also visible (to a lesser extent) in the $\tt{F500\_3E8}$ plot in Figure \ref{fig:compare_500Hz_abar}, with a faint haze of material rising above the flame. However, this is likely to be a numerical artefact rather than a true physical effect, and a result of the boundary conditions imposed at the axis of symmetry for these simulations.} $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}).

@@

The method for measuring flame propagation described in Section \ref{ssec:rot_propagation} is applied in Figure \ref{fig:flame_speeds_2} to the three $500~\mathrm{Hz}$ runs with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$. Due to the lack of a clear burning front in $\tt{F500\_4E8}$, we do not analyze its propagation speed and acceleration. As the initial $T_{\mathrm{star}}$ is increased, the flame becomes greatly accelerated. The initial flame velocities and accelerations derived from a quadratic least-squares fit to each of the datasets, {\color{blue} as well as the flame velocities at $t = 50~\mathrm{ms}$ calculated using the fit parameters,} are provided in Table~\ref{table:flame_speeds}. {\color{blue} Comparing the flame propagation at different initial temperatures, the most robust feature is the acceleration of the $\tt{F500\_3E8}$ run at $t \sim 40~\mathrm{ms}$.} The reason for the acceleration of the flame is not entirely clear. It could be that energy released from burning begins to dominate the flame's propagation as it evolves, increasing the flame speed over time. Another possibility is that the increased temperatures lead to enhanced turbulent mixing effects that pull in more fresh fuel for the flame to burn. Yet another possibility is that the higher initial $T_{\mathrm{star}}$ leads to a greater average temperature in the fuel layer over time, making it easier for the flame to burn the fuel and propagate.

>>>>

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. We saw in \citet{flame_wave1} that 
increasing the neutron star rotation rate reduces the horizontal lengthscale of the flame. An 
estimate of this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius 
may be thought of as the scale over which the balance between the Coriolis force and horizontal 
pressure gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use the energy generation rate $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region (where significant energy generation is occurring) as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). From Equation 
\ref{eqn:rossby}, we can see that increasing the rotation rate from $500~\mathrm{Hz}$ to 
$1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater confinement from 
the Coriolis force should reduce the horizontal extent of the flame by a similar factor. However, 
the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our 
simulations we see that this scaling does not work so well for all rotation rates. The simulations 
seem to follow the theoretical scaling described in Equation \ref{eqn:rossby} more closely at 
higher rotation rates ($1000~\mathrm{Hz}$ and higher), based on the results of \citet{flame_wave1}.

@@

The {\tt F500} and {\tt F1000} runs both qualitatively resemble the flame structure in 
\citet{flame_wave1} --- a laterally propagating flame that is lifted off of the bottom of the fuel 
layer --- but they differ in their burning structures. Figures~\ref{fig:time_series_500} and 
\ref{fig:time_series_1000} show time series of the mean molecular weight, $\bar{A}$, for the {\tt 
F500} and {\tt F1000} runs.  Compared to those in \cite{flame_wave1}, ashes behind the flame do not 
reach as high atomic weights. This is not surprising, since those previous runs artificially 
boosted the reaction rates.  Comparing these two new runs, the burning is much more evolved for the 
higher rotation rate, and the ash is actually able to move ahead of the flame front (visible in the 
Figure  \ref{fig:time_series_1000} $100~\mathrm{ms}$ snapshot). We believe that this is because the increased 
rotation better confines the initial perturbation and subsequent expansion from the burning, 
increasing the temperature and density in the flame front such that the reaction rate increases, 
which allows the reactions to progress further. The $\dot{e}_\mathrm{nuc}$ plots in Figure 
\ref{fig:time_series_enuc_1000} also support this interpretation, with the interior of the 
flame nearest to the crust in the {\tt F1000} run reaching higher $\dot{e}_\mathrm{nuc}$ values 
than for the {\tt F500} run in Figure \ref{fig:time_series_enuc_500}. In contrast to {\tt F500} 
and {\tt F1000}, the lowest rotation run --- {\tt F250} --- failed to ignite. The lack of ignition 
for {\tt F250} also aligns with the reasoning given above, with the lower rotation in this case 
potentially leading to insufficient confinement such that the temperature and density required for 
ignition is not achieved. In this scenario, another source of confinement (e.g.\ magnetic fields) 
would need to take over at lower rotation rates to allow a burst to occur, at least for the initial 
model used here.

@@

\begin{deluxetable}{lrrr}
	\tablecaption{\label{table:flame_speeds} Flame speed and acceleration values measured for each 
	simulation. $v_0$ is the flame velocity in the $+r$ direction at $t 
	= 0~\mathrm{ms}$. The initial flame 
	velocities and accelerations are derived from a quadratic least-squares fit to each of the 
	datasets for times $t \gtrsim 20~\mathrm{ms}$. Using these fit parameters we calculate the 
	velocity at $t = 50~\mathrm{ms}$, $v_{50}$, at which point the flames have reached steady 
	propagation.}
	\tablehead{\colhead{run} & \colhead{$v_0$ (km s$^{-1}$)} & \colhead{$a$ (km s$^{-2}$)} & \colhead{$v_{50}$ (km s$^{-1}$)}} 
	\startdata
  {\tt F1000}     & $3.414 \pm 0.008$ & $-1.03 \pm 0.13$ & $3.36 \pm 0.01$\\
  {\tt F500}      & $3.077 \pm 0.013$ & $5.41 \pm 0.19$ & $3.35 \pm 0.02$\\
  {\tt F500\_2E8} & $3.760 \pm 0.015$ & $1.74 \pm 0.25$ & $3.85 \pm 0.02$\\
  {\tt F500\_3E8} & $-5.293 \pm 0.100$ & $357.41 \pm 2.15$ & $12.58 \pm 0.15$\\
	\enddata
	%
\end{deluxetable}

@@

Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends 
across the domain and high into the atmosphere and fuel layer, lacking the characteristic flame 
structure shown in $\bar{A}$ plots for the lower temperature $500~\mathrm{Hz}$ runs (see 
Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material appears to have broken out of the 
atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 
10^4~\mathrm{cm}$. A similar effect is also visible (to a lesser extent) in the $\tt{F500\_3E8}$ 
plot in Figure \ref{fig:compare_500Hz_abar}, with a faint haze of material rising above the flame. 
However, this is likely to be a numerical artefact rather than a true physical effect, and a result 
of the boundary conditions imposed at the axis of symmetry for these simulations. $\tt{F500\_4E8}$ 
clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, 
$t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures 
\ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}).

@@

The method for measuring flame propagation described in Section \ref{ssec:rot_propagation} is 
applied in Figure \ref{fig:flame_speeds_2} to the three $500~\mathrm{Hz}$ runs with 
$T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$. Due to the lack of a clear burning front in 
$\tt{F500\_4E8}$, we do not analyze its propagation speed and acceleration. As the initial 
$T_{\mathrm{star}}$ is increased, the flame becomes greatly accelerated. The initial flame 
velocities and accelerations derived from a quadratic least-squares fit to each of the datasets, as 
well as the flame velocities at $t = 50~\mathrm{ms}$ calculated using the fit parameters, are 
provided in Table~\ref{table:flame_speeds}. Comparing the flame propagation at different initial 
temperatures, the most robust feature is the acceleration of the $\tt{F500\_3E8}$ run at $t \sim 
40~\mathrm{ms}$. The reason for the acceleration of the flame is not entirely clear. It could be 
that energy released from burning begins to dominate the flame's propagation as it evolves, 
increasing the flame speed over time. Another possibility is that the increased temperatures lead 
to enhanced turbulent mixing effects that pull in more fresh fuel for the flame to burn. Yet 
another possibility is that the higher initial $T_{\mathrm{star}}$ leads to a greater average 
temperature in the fuel layer over time, making it easier for the flame to burn the fuel and 
propagate.
